---
output: word_document
---
## Activity Monitoring Assignment

> Reproducible Research
> Peer Assessment 1
> annjinette

The following analyzes personal activity from a monitoring device for one individual for over 2 months, providing data at five minute intervals throughout.  We will look at 5 questions:  
  
*1. What is the mean total number of steps taken per day?  *
*2. What is the average daily activity pattern?  *
*3. Are there patterns in the missing values?  *
*4. Are there differences in activity patterns between weekends and weekdays?  *

### Loading and Preprocessing the Data
```{r}
##set working directory; create data folder if non exists
setwd("/Users/ann.hess/Desktop/files")
if(!file.exists("data")) {
        dir.create("data")
}
setwd("/Users/ann.hess/Desktop/files/data")

#read data
data <- read.csv("./activity 2.csv", header=T,
                      check.names=F, stringsAsFactors=F, comment.char="", quote='\"')
#format dates
data$date <- as.Date(data$date)
#using ggplot2 for visualizations
library(ggplot2)
```

### What is the mean total number of steps taken per day?  
#### Histogram - Total Steps  
```{r, echo=TRUE}
#aggregate data for analysis
totalSteps <- aggregate(steps ~ date, data = data, sum, na.rm=T)

#histogram total steps per day
qplot(steps, data = totalSteps, geom="histogram")  
```
#### Mean Steps Overall  
```{r, echo=TRUE}
mean(totalSteps$steps)
```
#### Median Steps Overall  
```{r, echo=TRUE}
median(totalSteps$steps)
```
### What is the average daily activity pattern?  
```{r, echo=TRUE}
#average activity pattern
avgSteps <- aggregate(steps ~ interval, data = data, mean, na.rm=T)
qplot(interval, steps, data = avgSteps)
```
### Are there patterns in the missing values?  
#### Number of Missing Values  
```{r, echo=TRUE}
sum(is.na(data))
```
#### Create Dataset with NA values replaces with mean in 5 minute interval  
```{r, echo=TRUE}
#replace NA values wiht mean in 5 min interval
StepsAverage <- aggregate(steps ~ interval, data = data, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(data)) {
        obs <- data[i, ]
        if (is.na(obs$steps)) {
                steps <- subset(StepsAverage, interval == obs$interval)$steps
        } else {
                steps <- obs$steps
        }
        fillNA <- c(fillNA, steps)
}
#new dataset
newData <- data
newData$steps <- fillNA
totalSteps2 <- aggregate(steps ~ date, data = newData, sum, na.rm=T)
```
#### Histogram - Total Steps by Day  
```{r, echo = TRUE}
qplot(steps, data = totalSteps2, geom="histogram")  
```
#### Mean Steps Overall - NAs Included  
```{r, echo = TRUE}
mean(totalSteps2$steps)
````
#### Median Steps Overall - NAs Included  
```{r, echo = TRUE}
median(totalSteps2$steps)
````
### Are there differences in activity patterns between weekends and weekdays?  
#### Create factor variable in dataset with 2 levels - weekend and weekday   
```{r, echo = TRUE}
day <- weekdays(data$date)
daylevel <- vector()
for (i in 1:nrow(data)) {
        if (day[i] == "Saturday") {
                daylevel[i] <- "Weekend"
        } else if (day[i] == "Sunday") {
                daylevel[i] <- "Weekend"
        } else {
                daylevel[i] <- "Weekday"
        }
}
data$daylevel <- daylevel
data$daylevel <- factor(data$daylevel)
````
#### Plot Activity Weekend vs Weekday   
```{r, echo = TRUE}
stepsByDay <- aggregate(steps ~ interval + daylevel, data = data, mean)
names(stepsByDay) <- c("interval", "daylevel", "steps")
qplot(interval, steps, data = stepsByDay, facets=.~daylevel)
````


